---
title: "Big Five Personality Traits"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The purpose of this project is to elevate the big five personality traits (Openness to experience, Conscientiousness, Extraversion, Agreeableness, Neuroticism) https://www.verywell.com/the-big-five-personality-dimensions-2795422. In relation to gender difference between males and females.

The second part of this project is to look at the differences between males and females and investigate whether there is a significant difference.

The third part of this project will look at the relationship between race, gender and personailty traits.

The fouth part of this project will look at the relationship between participants dominant hand, gender and personailty trait.

This data comes from the Open Source Psychometrics Project (https://openpsychometrics.org).

This dataset consists of 19720 participants (11985 females and 7608 males).

Raw file http://openpsychometrics.org/_rawdata/BIG5.zip

Website https://openpsychometrics.org/tests/IPIP-BFFM/

```{r echo=TRUE}
library(ggplot2)
library(car)
library(dplyr)
```

```{r}
set.seed(101)
pd <- read.delim("~/Downloads/BIG5/data.csv", header= TRUE)
pd <- na.omit(pd)
head(pd)
```

The scores for each of the five traits need to be summed into a new total score.

There are four different variables used for gender (1=Male, 2=Female, 3=Other 0=missed). This project is only looking at males and females, the other two variables (Other and missed) were removed using the subset function.

```{r}
pd$Extraversion  <- rowSums(pd[ ,8:17])
pd$Neuroticism  <- rowSums(pd[ ,18:27])
pd$Agreeableness  <- rowSums(pd[ ,28:37])
pd$Conscientiousness  <- rowSums(pd[ ,38:47])
pd$Openness  <- rowSums(pd[ ,48:57])
```

```{r}
female <- subset(pd, pd$gender == "2")
male <- subset(pd, pd$gender == "1")
pd <- rbind(female, male)
```

```{r}
pd$Sex <-car::recode(pd$gender, "1 ='M'; 2 ='F'")
```

##Gender difference in personality
These five density plots show personality differences between males and females. The blue overlay is male and the red overlay is female.

The lines on each graph represent the mean score (Males are blue and Females are red).
Note: higher or lower score on each item do not signify positive or negative traits.

###Openness to experience
Openness to experience is a trait associated with insight, imagination, creativity and adventure.

https://www.psychologistworld.com/influence-personality/openness-to-experience-trait

```{r echo=FALSE}
ggplot(data=pd, aes(x= Openness, fill= Sex)) + 
  geom_density(alpha=0.2, adjust=2) + 
  geom_vline(xintercept = mean(male$Openness), colour = "blue") +
  geom_vline(xintercept = mean(female$Openness), colour = "red") +
  xlab("Openness to experience") +
  ylab("Density") +
  ggtitle("Openness to experience") +
  theme(plot.title = element_text(hjust = 0.5))
```

###Conscientiousness
Conscientiousness is a personality trait associated with thoughtfulness, impulse control and goal-directed behaviour.

https://www.psychologistworld.com/influence-personality/conscientiousness-personality-trait

```{r echo=FALSE}
ggplot(data=pd, aes(x= Conscientiousness, fill= Sex)) + 
  geom_density(alpha=0.2, adjust=2) + 
  geom_vline(xintercept = mean(male$Conscientiousness), colour = "blue") +
  geom_vline(xintercept = mean(female$Conscientiousness), colour = "red") +
  xlab("Conscientiousness") +
  ylab("Density") +
  ggtitle("Conscientiousness") +
  theme(plot.title = element_text(hjust = 0.5))
```

###Extraversion
Extraversion is a personality trait that is associated with assertiveness, emotional, expression, talkative, sociable and excitability.

https://www.psychologistworld.com/influence-personality/extraversion-introversion

```{r echo=FALSE}
ggplot(data = pd, aes(x= Extraversion, fill= Sex)) + 
  geom_density(alpha=0.2, adjust=2) +
  geom_vline(xintercept = mean(male$Extraversion), colour = "blue") +
  geom_vline(xintercept = mean(female$Extraversion), colour = "red") +
  xlab("Extraversion") +
  ylab("Density") +
  ggtitle("Extraversion") +
  theme(plot.title = element_text(hjust = 0.5))
```

###Agreeableness
Agreeableness is a personality trait that is associated with trust, kindness and affection.

https://www.psychologistworld.com/personality/agreeableness-personality-trait

```{r echo=FALSE}
ggplot(data=pd, aes(x= Agreeableness, fill= Sex)) + 
  geom_density(alpha=0.2, adjust=2) + 
  geom_vline(xintercept=mean(male$Agreeableness), colour = "blue") +
  geom_vline(xintercept=mean(female$Agreeableness), colour = "red") +
  xlab("Agreeableness") +
  ylab("Density") +
  ggtitle("Agreeableness") +
  theme(plot.title = element_text(hjust = 0.5))
```

###Neuroticism
Neuroticism is a personality trait that is associated with emotional instability, anxiety, irritability and sadness.

https://www.psychologistworld.com/personality/neuroticism-personality-trait

```{r echo=FALSE}
ggplot(data = pd, aes(x= Neuroticism, fill = Sex)) +
  geom_density(alpha=0.2, adjust=2) +
  geom_vline(xintercept = mean(male$Neuroticism), colour = "blue") +
  geom_vline(xintercept = mean(female$Neuroticism), colour = "red") +
  xlab("Neuroticism") +
  ylab("Density") +
  ggtitle("Neuroticism") +
  theme(plot.title = element_text(hjust = 0.5))
```

This data shows that female had a higher average score for Conscientiousness, agreeableness and Neuroticism. Whereas, males had a higher average score for Openness to experience and Extraversion.

##Race & Personality Traits

The next part of this project looks at the relationship between race, gender and personality trait. The dataset recorded 9 different races were collected Arctic (Siberian, Eskimo), Caucasian(European, Indian, Middle East, North African), Indigenous Australian, Native American, North East Asian (Mongol, Tibetan, Korean Japanese), Pacific (Polynesian, Micronesian), South East Asian (Chinese, Thai, Malay, Filipino), West African, Bushmen, Ethiopian. Two further races where recored Mixed race and Other, finally, some participants choose to not included their race recorded as missing.

```{r echo=FALSE}
pd$race <- recode_factor(pd$race, '0' = "missed",
                         '1' = "Mixed Race",
                         '2' = "Arctic (Siberian, Eskimo)",
                         '3' = "Caucasian (European)",
                         '4' = "Caucasian (Indian)",
                         '5' = "Caucasian (Middle East)",
                         '6' = "Caucasian (North African, Other)",
                         '7' = "Indigenous Australian",
                         '8' = "Native American",
                         '9' = "North East Asian (Mongol, Tibetan, Korean, Japanese)",
                         '10' = "Pacific (Polynesian, Micronesian)",
                         '11' = "South East Asian (Chinese, Thai, Malay, Filipino)",
                         '12' = "West African, Bushmen, Ethiopian",
                         '13' = "Other")
```

###Extraversion

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(race), y = Extraversion)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Extraversion") +
  ylab("Race") +
  ggtitle("Extraversion") +
  theme(plot.title = element_text(hjust = 0.5))
```
This Barplot indicates that there is little extraversion scores for each of the races. However, based on the scores Caucasian (European) have the lowest levels of extraversion with a mean score of 30.38. Whereas, Arctic (Siberian, Eskimo) scored the highest for extraversion with a mean score of 33.15.


###Neuroticism

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(race), y = Neuroticism)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) + 
  xlab("Neuroticism") +
  ylab("Race") +
  ggtitle("Neuroticism") +
  theme(plot.title = element_text(hjust = 0.5))
```

This boxplot indicates that Arctic (Siberian, Eskimo) have the lowest levels of Neuroticism with a mean of 27.46, this would suggest that Arctic (Siberian, Eskimo) have lower levels of anxiety and depression (https://www.psychologytoday.com/us/basics/neuroticism). However, other data suggests that other white Caucasian, Eskimo where the second most lilkey to use mental health services (https://www.samhsa.gov/data/sites/default/files/MHServicesUseAmongAdults/MHServicesUseAmongAdults.pdf)(P1). Whereas, Caucasian (Indian) have the highest levels of Neuroticism with a mean of 32.58.

###Agreeableness

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(race), y = Agreeableness)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Agreeableness") +
  ylab("Race") +
  ggtitle("Agreeableness") +
  theme(plot.title = element_text(hjust = 0.5))
```
This data indicates that Caucasian (European) have the lowest levels of Agreeableness (31.53). Whereas, Arctic (Siberian, Eskimo) have the highest levels of Agreeableness (34.30)

###Conscientiousness

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(race), y = Conscientiousness)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Conscientiousness") +
  ylab("Race") +
  ggtitle("Conscientiousness") +
  theme(plot.title = element_text(hjust = 0.5))
```

The boxplot would suggest that there is very little difference between races interms of Conscientiousness. Although, Indigenous Australian (29.68) scored the highest, Whereas, South East Asian (Chinese, Thai, Malay, Filipino) (31.88) scored the highest on the Conscientiousness trait.

###Openness to experience

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(race), y = Openness)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Openness") +
  ylab("Race") +
  ggtitle("Openness") +
  theme(plot.title = element_text(hjust = 0.5))
```

The final Trait Openness to experience indicates that there is a significant amount of overlap between races. However, Caucasian (Middle East) scored the lowest (32.25). Whereas, Caucasian (European) scored the highest (33.65) for Openness to experience. 


##Personailty Traits & hand preference
This examines the differences between males and females for each of the personality traits in relation to hand preference.

```{r}
pd$hand <- recode_factor(pd$hand, '0' = "missed",
                         '1' = "Right",
                         '2' = "Left",
                         '3' = "Both")
```

###Extraversion

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(hand), y = Extraversion, col = Sex)) +
  geom_boxplot() +
  xlab("Extraversion") +
  ylab("Race") +
  ggtitle("Extraversion") +
  theme(plot.title = element_text(hjust = 0.5))
```
The first boxplot indicates that males scored higher then females for Extraversion. Furthermore, both males and females that had no prefered hand had higher levels of extraversion compared to individuals that had a prefered hand.

###Neuroticism

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(hand), y = Neuroticism, col = Sex)) +
  geom_boxplot() +
  xlab("Neuroticism") +
  ylab("Race") +
  ggtitle("Neuroticism") +
  theme(plot.title = element_text(hjust = 0.5))
```
This boxplot indicates that females scored higher then males for Neuroticism. Although, there was no difference between males and females for individuals that had no prefered hand. Furthermore, females that had no prefer hand had the lowest levels of Neuroticism.

###Agreeableness

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(hand), y = Agreeableness, col = Sex)) +
  geom_boxplot() +
  xlab("Agreeableness") +
  ylab("Race") +
  ggtitle("Agreeableness") +
  theme(plot.title = element_text(hjust = 0.5))
```
This boxplot indicates that left handed males are the most Agreeableness (29.92) where as left handed females have the lowest levels of Agreeableness (32.09). Furthermore, females had higher levels of Agreeableness compared to males. 

###Conscientiousness

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(hand), y = Conscientiousness, col = Sex)) +
  geom_boxplot() +
  xlab("Conscientiousness") +
  ylab("Race") +
  ggtitle("Conscientiousness") +
  theme(plot.title = element_text(hjust = 0.5))
```
This boxplot indicates that both males and females with no hand preference scored higher compared to the other groups. Whereas, right handed individuals scored the lowest for Conscientiousness.

###Openness to experience

```{r echo=FALSE}
ggplot(data = pd, aes(x = factor(hand), y = Openness, col = Sex)) +
  geom_boxplot() +
  xlab("Openness") +
  ylab("Race") +
  ggtitle("Openness") +
  theme(plot.title = element_text(hjust = 0.5))
```
This boxplot indicates that both males and females that had no hand preference scored higher for Openness to experience compared to the other groups, Furthermore there was much overlap between the other three groups.

##Analysis
The second part of this project is to investigate whether the difference displayed in the plots above are significant, Therefore, is there a significant difference between Male and Female personality types?

```{r}
model <- manova(cbind(Openness, Conscientiousness, Extraversion, Agreeableness, Neuroticism) ~ gender, data = pd)
summary.aov(model)
```

This MANOVA model shows that there is a significant difference for each of the five personality types between males and females. This model would indicate that the most significant differences are Extraversion and Neuroticism.

###Conclusion